---
title: "Education Attainment in UK"
output: 
  html_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(gtsummary)
library(pwr)
library(knitr)

eng_ed <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-01-23/english_education.csv')

```



## 1) Data Cleaning

| **Variable**                                             | **Reason for Selection**                                                                                         |
|----------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------|
| `town11nm`                                               | Identifies the geographical area, essential for studying regional patterns.                                      |
| `size_flag`                                              | Captures town size, a key variable for analyzing differences in educational outcomes.                            |
| `income_flag`                                            | Represents income levels, crucial for examining socioeconomic influences on education.                           |
| `uni_flag`                                        | Indicates if a town has a university, potentially impacting higher education access.                             |
| ` qual_residents`                         | Reflects the proportion of educated adults, a possible community influence on students.                          |
| `GCSEs`        | Serves as a proxy for high school graduation rates.                                                              |
| `college_grad` | Represents college graduation rates.                                                                             |


| **Exclusion Criteria**                                                                                       |
|-------------------------------------------------------------------------------------------------------------|
| - Dropped variables not directly related to the research question, such as minor demographic details or redundant information. |
| - Variables with incomplete or irrelevant data for analyzing educational attainment (e.g., administrative or non-educational metrics).  |


```{r}
# Selecting relevant columns and renaming variables for clarity
eng_ed <- eng_ed |> 
  select(town11nm, size_flag, income_flag, university_flag,
         level4qual_residents35_64_2011,
         key_stage_4_attainment_school_year_2012_to_2013,
         highest_level_qualification_achieved_by_age_22_level_6_or_above,
         ) |> 
    mutate(
    size_flag = factor(size_flag),
    income_flag = factor(income_flag),
    uni_flag= factor(university_flag),
    qual_residents=factor(level4qual_residents35_64_2011),
    GCSEs=key_stage_4_attainment_school_year_2012_to_2013,
    college_grad =  
      highest_level_qualification_achieved_by_age_22_level_6_or_above
   ) 

#removing repeat columns
eng_ed <- eng_ed |> 
  select(town11nm, size_flag, income_flag, uni_flag,qual_residents, 
                 GCSEs, college_grad   )

```

```{r, echo=F, include=F}
glimpse(eng_ed)
summary(eng_ed)
```

**Reason for Collapsing Town Size Categories**

| **Collapsed Level**  | **Original Categories Included**                                                                                         | **Reason**                                                                                                   |
|-----------------------|--------------------------------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------|
| `Large`              | "City", "Medium Towns", "Large Towns", "Outer London BUA", "Inner London BUA", "Not BUA"                                  | Reflects similarities in urban characteristics, such as population density and access to resources.         |
| `Small`              | "Small Towns", "Other Small BUAs"                                                                                        | Focuses on smaller, less urbanized communities to highlight contrasts with larger urban areas.              |

**Additional Notes**

- **Simplification Benefits:** Collapsing categories reduces noise and ensures clearer distinctions between urban and rural-like areas.  
- **Comparison Focus:** The new levels, `Large` and `Small`, facilitate easier interpretation of differences in educational attainment based on town size.


```{r}
#factor collapse to combine mid and large size towns and cities into one level
levels(eng_ed$size_flag)
eng_ed<- eng_ed |>
mutate(size_flag = fct_collapse(size_flag,
Large = c("City", "Medium Towns", "Large Towns", "Outer london BUA", 
          "Inner London BUA", "Not BUA"),
Small = c("Small Towns",  "Other Small BUAs")
)) 
```


## 2) EDA
```{r}
 #distribution of the size_flag
ggplot(data = eng_ed, aes(x = size_flag, fill = size_flag)) + 
  geom_bar() + 
  scale_fill_brewer(palette = "Pastel1") + 
  labs(
    title = "Distribution of Town Sizes",
    x = "Town Size",
    y = "Frequency",
    fill = "Town Size"  
  ) + 
  theme_minimal(base_size = 10) + 
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top" 
  )

```

```{r}
#distribution of the income_flag
eng_ed |> 
  drop_na() |>  
  ggplot(aes(y = income_flag, fill = income_flag)) + 
  geom_bar() + 
  scale_fill_brewer(palette = "Pastel1") + 
  labs(
    title = "Distribution of Income Status of the Towns",
    y = "Income Status",
    x = "Frequency",
    fill = "Income Status"
  ) + 
  theme_minimal(base_size = 10) + 
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5), 
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top"  
  )
```

```{r}
# Distribution of High School Graduation
ggplot(eng_ed, aes(x = GCSEs)) + 
  geom_histogram(
    bins = 10,  
    col = "white", 
    fill = "#DECBE4"
  ) + 
  labs(
    title = "Distribution of High School Graduation",
    x = "High School Graduation Proportion of Towns",
    y = "Frequency"
  ) + 
  theme_minimal(base_size = 10) +  
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),  
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

```

```{r}
# Distribution of College Graduation
ggplot(eng_ed, aes(x = college_grad)) + 
  geom_histogram(
    bins = 10,       
    col = "white",    
    fill = "#CCEBC5" 
  ) + 
  labs(
    title = "Distribution of College Graduation",
    x = "College Graduation Proportion of Towns",
    y = "Frequency"
  ) + 
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )


```



```{r, results = "asis"}
# Summary Statistics Grouped by Town Size
eng_ed |> 
  select(size_flag, income_flag, GCSEs, college_grad) |> 
  tbl_summary(
    by = size_flag,
     # Display mean (SD) for continuous variables
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    # Treat dichotomous variables as categorical
    type = list(all_dichotomous() ~ "categorical"),  
    # Round to 2 decimal places
    digits = list(all_continuous() ~ c(2, 2)),            
    label = list(
      income_flag = "Income Flag",
      GCSEs = "High School Graduation",
      college_grad = "College Graduation"
    )
  ) |> 
  as_kable() 

```

## 3) Data Analysis

**TEST 1: High School Graduation Rates: Small v Large Towns**

Is there a difference in high school graduation rates between small v large towns/cities?

- H0: There is no mean difference between the high school graduation rates  between small v large towns/cities
- H1: There is a difference  between the high school graduation rates  between small v large towns/cities
- Test: Difference between two means

**Conditions**

- *Independence*: We can assume the observations of one town to another are independent from each other.
- *Normality*: Populations are approximately normal
- *Sample Size*: Both groups have n larger than 30



```{r}
small_town <- eng_ed |> 
  filter(size_flag == "Small") #filtered data by small town

large_town <- eng_ed |> 
  filter(size_flag == "Large") #filtered data by large town


nrow(small_town)
nrow(large_town)

```


**Power Analysis**

```{r}
mean_small_gcse <- mean(small_town$GCSEs, na.rm = TRUE)
mean_large_gcse <- mean(large_town$GCSEs, na.rm = TRUE)

sd_pooled_gcse <- sqrt((
  (sd(small_town$GCSEs, na.rm = TRUE)^2 + 
     sd(large_town$GCSEs, na.rm = TRUE)^2) / 2
))

# Cohen's d (effect size)
cohens_d_gcse <- abs(mean_small_gcse - mean_large_gcse) / sd_pooled_gcse

# Sample size
n_small <- nrow(small_town)
n_large <- nrow(large_town)

# Power analysis
power_gcse <- pwr.t.test(
  d = cohens_d_gcse,   
  n = min(n_small, n_large), 
  sig.level = 0.05,   
  type = "two.sample" 
)$power

cat(
  "Results Summary for GCSE Difference Test:\n",
  "------------------------------------------\n",
  "Observed Cohen's d (Effect Size):", round(cohens_d_gcse, 3), "\n",
  "Sample Size (Small Towns):", n_small, "\n",
  "Sample Size (Large Towns):", n_large, "\n",
  "Computed Power for GCSE Difference Test Power Analysis", round(power_gcse, 3), "\n"
)

```


**Test for Difference in Means**

```{r}
t_test_result <- t.test(small_town$GCSEs, large_town$GCSEs)
cat("T-test Result: \n")
print(t_test_result)
```

- Decision: The p-value is less than 0.05, we reject the null hypothesis
- Conclusion: We have enough evidence that there is a significant difference in high school graduation rates between small towns and large towns/cities.

_______________________________

**TEST 2: Difference in College Graduation Rates: Small v Large Towns**

Is there a difference in college graduation rates between small  v large towns/cities?

- H0: There is no difference in college graduation rates between small vs large towns/cities.
- H1: There is a difference in college graduation rate between small vs large towns/cities.
- Test: Difference between two means

**Conditions**

- *Independence*: We can assume the observations of one town to another are independent from each other.
- *Normality*: Populations are approximately normal
- *Sample Size*: Both groups have n larger than 30


**Power Analysis**

```{r}
mean_small_college <- mean(small_town$college_grad, na.rm = TRUE)
mean_large_college <- mean(large_town$college_grad, na.rm = TRUE)

sd_pooled_college <- sqrt((
  (sd(small_town$college_grad, na.rm = TRUE)^2 + 
     sd(large_town$college_grad, na.rm = TRUE)^2) / 2
))

# Cohen's d (effect size)
cohens_d_college <- abs(mean_small_college - mean_large_college) / 
  sd_pooled_college

# Power analysis for two-sample t-test
power_college <- pwr.t.test(
  d = cohens_d_college,   
  n = min(n_small, n_large), 
  sig.level = 0.05,       
  type = "two.sample"     
)$power


cat(
  "Results Summary for College Graduation Difference Test:\n",
  "--------------------------------------------------------\n",
  "Observed Cohen's d (Effect Size):", round(cohens_d_college, 3), "\n",
  "Sample Size (Small Towns):", n_small, "\n",
  "Sample Size (Large Towns):", n_large, "\n",
  "Computed Power for College Graduation Difference Test:", round(power_college, 3), "\n"
)


```


**Test for Difference in Means**

```{r}
t_test_college_grad <- t.test(small_town$college_grad, large_town$college_grad)
cat("T-test Results for College Graduation Rates: \n")
print(t_test_college_grad)
```

- Decision: The p-value is less than 0.05, we reject the null hypothesis
- Conclusion: We enough evidence that there is a difference in college graduation rate between small vs large towns/cities.

_______________________________

**TEST 3: Association between Income and Town Size **

Are income and town sizes associated?

- H0: Income levels are independent of town size.
- H1: Income levels are associated with town size
- Test: Chi-squared test of independence

**Conditions**

- *Independence*: We can assume the observations of one town to another are independent from each other.
- *Expected Counts*: All counts are greater than 5. 

```{r}
income_size_table <- table(eng_ed$size_flag, eng_ed$income_flag)
proportions <- prop.table(income_size_table, margin = 1)

proportions_df <- as.data.frame(as.table(proportions))
colnames(proportions_df) <- c("Town_Size", "Income_Level", "Proportion")

# Bar chart
ggplot(proportions_df, aes(x = Income_Level, y = Proportion, 
                           fill = Town_Size)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), col = "white") + 
  labs(
    title = "Proportion of Towns by Income Level and Town Size",
    x = "Income Level",
    y = "Proportion of Towns",
    fill = "Town Size"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +  
  scale_fill_manual(
    values = c("Small" = "#B3CDE3", "Large" = "#FBB4AE") 
  ) +
  theme_minimal(base_size = 10) +  # Consistent font size
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5), 
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )

```

**Power Analysis**

```{r}
chisq_test <- chisq.test(eng_ed$size_flag, eng_ed$income_flag)

income_size_table <- table(eng_ed$size_flag, eng_ed$income_flag)

# Calculate Cramér's V
cramers_v <- sqrt(chisq_test$statistic / (sum(income_size_table) * 
                                            (min(dim(income_size_table)) - 1)))


# Df
df <- chisq_test$parameter

# Total sample size
n_total <- sum(income_size_table)

# Compute power for chi-squared test
power_chisq <- pwr.chisq.test(
  w = cramers_v,  
  N = n_total,  
  df = df,       
  sig.level = 0.05  
)$power


cat(
  "Results Summary for Chi-Squared Test:\n",
  "-------------------------------------\n",
  "Observed Cramér's V (Effect Size):", round(cramers_v, 3), "\n",
  "Degrees of Freedom:", df, "\n",
  "Total Sample Size:", n_total, "\n",
  "Computed Power for Chi-Squared Test:", round(power_chisq, 3), "\n"
)

```


```{r}
test <- chisq.test(eng_ed$income_flag, eng_ed$size_flag)
cat("Expected counts:", test$expected)
test

```
- Decision: We reject the null hypothesis.
- Conclusion: There is strong statistical evidence to conclude that income levels and town size are not independent. 

_______________________________

**TEST 4: Relationship between College Graduation Rate & Highschool Graduation Rate**

Is high school completion a good predictor of college degree completion?  

- H0: There is no linear relationship between high school and college completion.
- H1: There is a linear relationship between high school and college completion.
- Test: Linear Regression

**Conditions**

- *Linearity*: Data appears to be linear
- *Independence:* Errors seem to have no pattern
- *Constant Variance*: Seems Constant
- *Normality*: The qq plots appear normal

```{r}

#Independence

mod<- lm(college_grad ~ GCSEs, data =eng_ed)
plot(predict(mod), rstandard(mod))
abline(h=0)
summary(mod)

#constant variance
#normality 
plot(mod, 1:2)
```

## Power Analysis

```{r}
r_squared <- summary(mod)$r.squared

# Size (Cohen's f2)
f2 <- r_squared / (1 - r_squared)

n <- nrow(eng_ed) 
num_predictors <- 1

# Power analysis 
power_regression <- pwr.f2.test(
  u = num_predictors, 
  v = n - num_predictors - 1,
  f2 = f2,  
  sig.level = 0.05 
)$power

cat(
  "Results Summary for Regression Test:\n",
  "-------------------------------------\n",
  "R-squared (Goodness-of-Fit):", round(r_squared, 3), "\n",
  "Effect Size (Cohen's f²):", round(f2, 3), "\n",
  "Sample Size:", n, "\n",
  "Computed Power for Regression Test:", round(power_regression, 3), "\n"
)

```

```{r}
clean_data <- eng_ed[!is.na(eng_ed$GCSEs) & !is.na(eng_ed$college_grad), ]

# Fit the linear regression model
model <- lm(college_grad ~ GCSEs, data = clean_data)

# Calculate Adjusted R-squared
adj_r2 <- summary(model)$adj.r.squared

# Create scatter plot with regression line
ggplot(clean_data, aes(x = GCSEs, y = college_grad)) +
  geom_point(color = "steelblue", alpha = 0.6) + 
  geom_smooth(method = "lm", color = "red", se = TRUE) +  
  annotate("text", x = 40, y = max(clean_data$college_grad) - 5,
           label = paste("Adjusted R² = ", round(adj_r2, 3)),
           color = "darkred", size = 4, hjust = 0) + 
  labs(
    title = "Relationship Between High School and College Graduation Rates",
    x = "High School Graduation Rate (GCSEs)",
    y = "College Graduation Rate (%)"
  ) +
  theme_minimal()

```

- Decision: We reject the null hypothesis.
- Conclusion: The results provide strong evidence of a statistically significant positive linear relationship between high school graduation rates (`GCSEs`) and college graduation rates (`college_grad`). This indicates that high school completion is a good predictor of college degree completion. Specifically, the model estimates that for every 1% increase in high school graduation rates, college graduation rates increase by approximately 0.682%, on average. The model explains 59.8% of the variability in college graduation rates (\(R^2 = 0.598\)).




